CTL mapping - The added value from segregating correlation

### 2012-01-31 inst/methodPaper/document.tex
Paper updates
Code for paper figures

\begin{document}
\firstpage{1}
\title[CTLmapping]{CTL mapping - The added value from segregating correlation}
\author[Arends \emph{et al.}]{
  Danny Arends\,$^{1,\footnote{to whom correspondence should be addressed}}$,
  and Ritsert C. Jansen\,$^{1}$
}
\address{
  $^{1}$Groningen Bioinformatics Centre, University of Groningen, Groningen, The Netherlands.
}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXXXX}
\maketitle
\begin{abstract}
  \section{Motivation:}
  Here we present Correlated Traits Locus (CTL) mapping, a novel approach 
  to detect genetic regulation of traits in natural and inbred populations. 
  CTL mapping allows association of trait correlation differences to genetic 
  markers, in almost the same way as QTL mapping associates genetic  markers with differences in expression.

@@

  \section{Results:}
  The software is available under the terms of GNU General Public License, version 3
  (\href{http://www.r-project.org/Licenses/GPL-3}{www.r-project.org/Licenses/GPL-3}).
  \section{Availability:}
  Software, documentation and source code are available from:
  \href{http://www.mapctl.org}{www.mapctl.org}
  \section{Contact:} 
  \href{Danny.Arends@Gmail.com}{Danny.Arends@Gmail.com}
\end{abstract}

\section{Introduction}
  Genetical genomics experiments have shown that gene expression variation can be mapped to 
  genotypic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
  gene identifies regions in the genome for which different genotypes lead to differential 
  expression level of the gene. This QTL mapping can be done on all molecular levels including, 
  gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).

  We present Correlated Traits Locus (CTL) mapping as a comparable method to QCL mapping.
  CTL unlike QTL links differences in correlation to genetic variation, i.e. to identify regions in the 
  genome for which one genotype value leads to correlated expression between a trait and  
  other traits, while the other genotype doesn't show this correlation. CTL exploits the fact that 
  any QTL will be the result of either direct induced variation or indirect variation. The latter is 
  visible as correlation differences among individuals carrying different genotypes.

@@

  \begin{figure*}[ht]
    \begin{center}
    \vspace{-0.2cm}
    \hspace{-1.3cm}
    \subfigure[CTL profiles]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.pdf}
      \label{subfiga}
    }
    \hspace{0.3cm}
    \subfigure[Traits x Traits interaction matrix]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.pdf}
      \label{subfigb}
    }
    \hspace{0.3cm}
    \subfigure[Reconstructed network]{
      \includegraphics[keepaspectratio,scale=.2]{img/ctl_net.pdf}
      \label{subfigc}
    }
  \end{center}
  \begin{minipage}{7in}
  \label{newplots}
  \vspace{-0.7cm}
  \caption[Plots]{
  {\emph {\bf a}}: Example CTL + QTL profiles. Seen are the QTL profile on top of the CTL profile. 
  The first peak is a CTL, we don't observe a QTL for this location. The second peak is a QTL peek, with no underlying CTL. 
  {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
  showing for each trait the significance of the difference in correlation to all other traits. This 
  matrix can be transformed into .sif file to generate a network image. 
  {\emph {\bf c}}:  Reconstructed network, this shows the same information as the TxT matrix only in a more accessible way, we visualized our 
  network using cytoscape, the linewidth is the summarized interaction evidence.
 }
\end{minipage}
\vspace{-0.5cm}
\end{figure*}

@@

\emph{ {\bf Materials}}
  Metabolite data from an \emph{A. Thaliana} Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
  dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
  There are 162 recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
  details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}
\vspace{-0.6cm}
\section{Results}
  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was 
  determined by using 15,000 permutations.
  When combining CTL and QTL information, we observe:\\
  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - Peak 1). Loss of variation in the genes 
  underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
  detect the differences in correlation.\\
  2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
  are contributing variation to the observed QTL  (Fig \ref{subfiga} - Peak 3).\\
  3) QTL with no CTL  (Fig \ref{subfiga} - Peak 2), we interpret this as the source of the variation. Analysis in this way of 
  all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
  source of natural variation and follow it down the network.

>>>>

\begin{document}
\firstpage{1}
\title[CTLmapping]{CTL mapping - The added value from segregating correlation}
\author[Arends \emph{et al.}]{
  Danny Arends\,$^{1,\footnote{to whom correspondence should be addressed}}$,
  and Ritsert C. Jansen\,$^{1}$
}
\address{
  $^{1}$Groningen Bio-informatics Centre, University of Groningen, Groningen, The Netherlands.
}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXX}
\maketitle
\begin{abstract}
  \section{Motivation:}
  Here we present Correlated Traits Locus (CTL) mapping, a novel approach 
  to detect genetic regulation of traits in natural and inbred populations. 
  CTL mapping allows association of trait correlation differences to genetic 
  markers, in almost the same way as QTL mapping associates genetic  markers with differences in expression.

@@

  \section{Results:}
  The software is available under the terms of GNU General Public License, version 3
  (\href{http://www.r-project.org/Licenses/GPL-3}{www.r-project.org/Licenses/GPL-3}).
  \section{Availability:}
  Software, documentation and source code are available at request.
  \section{Contact:} 
  \href{Danny.Arends@Gmail.com}{Danny.Arends@Gmail.com}
\end{abstract}

\section{Introduction}
  Genetical genomics experiments have shown that gene expression variation can be mapped to 
  genetic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
  gene identifies regions in the genome for which different genotypes lead to differential 
  expression level of the gene. This QTL mapping can be done on all molecular levels including, 
  gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).

  We present Correlated Traits Locus (CTL) mapping as a comparable method to QTL mapping.
  CTL unlike QTL links differences in correlation to genetic variation, i.e. to identify regions in the 
  genome for which one genotype value leads to correlated expression between a trait and  
  other traits, while the other genotype doesn't show this correlation. CTL exploits the fact that 
  any QTL will be the result of either direct induced variation or indirect variation. The latter is 
  visible as correlation differences among individuals carrying different genotypes.

@@

  \begin{figure*}[ht]
    \begin{center}
    \vspace{-0.2cm}
    \hspace{-1.3cm}
    \subfigure[CTL profiles]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.pdf}
      \label{subfiga}
    }
    \hspace{0.3cm}
    \subfigure[Traits x Traits interaction matrix]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.pdf}
      \label{subfigb}
    }
    \hspace{0.3cm}
    \subfigure[Reconstructed network]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_net.pdf}
      \label{subfigc}
    }
  \end{center}
  \begin{minipage}{7in}
  \label{newplots}
  \vspace{-0.7cm}
  \caption[Plots]{
  {\emph {\bf a}}: Example CTL + QTL profiles. Seen are the QTL profile on top of the CTL profile. 
  The first peak is a CTL, we don't observe a QTL for this location. The second peak is a QTL peek, with no underlying CTL. 
  {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
  showing for each trait the significance of the difference in correlation to all other traits. This 
  matrix can be transformed into .sif file to generate a network image. 
  {\emph {\bf c}}:  Reconstructed network, this shows the same information as the TxT matrix only in a more accessible way, we visualized our 
  network using cytoscape, the width of the line is the maximum interaction evidence.
 }
\end{minipage}
\vspace{-0.5cm}
\end{figure*}

@@

\emph{ {\bf Materials}}
  Metabolite data from an \emph{A. Thaliana} Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
  dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
  There are 162 recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
  details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}
\vspace{-0.6cm}
\section{Results}
  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was 
  determined by using 15,000 permutations.
  When combining CTL and QTL information, we observe:\\
  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - P1). Loss of variation in the genes 
  underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
  detect the differences in correlation.\\
  2) QTL with no CTL (Fig \ref{subfiga} - P2), we interpret this as the source of the variation. Analysis in this way of 
  all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
  source of natural variation and follow it down the network.
  3) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
  are contributing variation to the observed QTL  (Fig \ref{subfiga} - P3).\\
